---
title: 'Intro to R: Day 1'
author: "Kumar Ramanathan"
date: "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation:
    theme: moon
    incremental: true
    self_contained: false
    reveal_plugins: ["notes"]
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
notes<-params$notes
```

# Introduction

## Why R?

## RStudio and RStudio Server

## Some key concepts

## Recall your DataCamp course

By now, you should all have completed the *Introduction to R* course on DataCamp. Congratulations! You already know quite a lot about how R works. 

To refresh, you covered the following material in your course:

- Arithmetic with R
- Objects and variable assignment
- Data types, including: character, numeric, logical, factors
- Data structures, including: vectors, matricies, data frames, lists

<aside class="notes">
Today, we will briefly review the above topics, and then turn to new material on reading and writing files, data manipulation, and data visualization.
</aside>

# Review: Basics and data types 

## Arithmetic

Open up the file `day1_exercises.R` in RStudio. Do the following tasks:

- Pick a number; save it as `x`
- Multiply x by 3
- Take the log of the above <small>(Hint, you need the function `log()` here)</small>
- Subtract 4 from the above
- Square the above

## Arithmetic

```{r arithmetic}
x <- 5
x*3
log(x*3)
log(x*3)-4
(log(x*3)-4)^2
```

## Comparisons and Logical Operators

- Check if 1 is bigger than 2
- Check if 1 + 1 is equal to 2
- Check if it is true that the strings "eat" and "drink" are not equal to each other
- Check if it is true that 1 is equal to 1 *AND* 1 is equal to 2 
  <small>(Hint: remember what the operators `&` and `|` do)</small>
- Check if it is true that 1 is equal to 1 *OR* 1 is equal to 2

## Comparisons and Logical Operators

```{r comparisons}
1 > 2
(1 + 1) == 2
"eat" != 'drink'
(1==1) & (1==2)
(1==1) | (1==2)
```

## Packages

- Install the package `tidyverse`
- Load the package `tidyverse`
- Open the help file for the function `recode` 
  <small>(Hint: remember what `?` does)</small>

## Packages

```{r packages, eval=F}
install.packages("tidyverse")

library(tidyverse)

?recode
```

## Data types

- **logical**: `TRUE` or `FALSE` 
  <small>(When co-coerced into numeric type, `TRUE`=`1` and `FALSE`=`0`)</small>
- **integer**: a specific type; most numbers are numeric instead
- **numeric**: real or decimal 
- **complex**: ex: `2+5i`
- **character**: `"text data"`, denoted with quotes
  <small>(`'single quotes'` and `"double quotes"` both work)</small>
- Use `typeof()` on an object to check its type

# Review: Data structures

## Vectors

## Factors

## Lists

## Matricies

## Data frames

# Reading files

# Data manipulation

# Writing files

# Data visualization